---
title: "Should we postpone our July 2020 wedding?"
author: "Isabelle Feldhaus, Antoine Lizée"
date: "2020-05-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', fig.width=7, fig.height=6, echo=FALSE, message=FALSE)
source("0. analysis.R")
```

With the World Health Organization's declaration of the Covid-19 pandemic in March 2020, the looming question has been whether or not to postpone our July wedding. A difficult decision before us, we wanted to take a quick look at the numbers to give some confidence to our decision, while keeping in mind that there are two key dimensions to consider: (1) health and (2) geopolitics. 

To be clear, we're not modeling or projecting anything here. We only wanted to look at the data ourselves since we couldn't find our questions answered out there[^1]. Since we figured that we're not the only nerds on this email list, we stitched the graphs together for your entertainment.

[^1]: This is also our coping mechanism. 

We used data collated by the [New York Times](https://github.com/nytimes/covid-19-data/raw/master/us-states.csv) (yes, NYT appears to be the cleanest and most complete data source at the moment) and the [European CDC](https://opendata.ecdc.europa.eu/covid19/casedistribution/csv) from mid-February to May 3rd to get a picture of Covid-19 trajectories in the United States and selected European countries, respectively. We computed the cumulative cases and deaths over the course of the pandemic, incidence (cases per day), and mortality (deaths per day) and adjusted for population size for each state or country. 

Let's take a look. 

### The (Not So) United States

Indeed, there have been a lot of deaths in recent weeks...

```{r deaths_per_day}
plot(deaths_per_day)
```

This graph shows the number of deaths per day in the US. For comparison, at its peak, Covid-19 caused more than **2,000** fatalities per day when in normal times, we'd expect about **7,700** deaths each day on average. That's over 25%! 

Note the dips in the graph. These are weekends and, subsequently, pauses in data collection. Because of these idiosyncracies in reporting, all the following figures show values for the past 7 days. 

```{r us_incidence}
us_incidence
```

So, this graph is a _mess_. But we see a few interesting trends. 

New York and New Jersey are moving together. Massachusetts is coming to get them. All that is unsurprising given their geographical proximity and interconnections. 

We also see plateauing. We can surmise that this mirrors the stay-home orders in various states. Good examples are Washington, New York, and Michigan (good luck finding this one). 

```{r us_deaths }
us_deaths
```

Looking at cases alone is not enough: case reporting is highly variable depending on testing protocols and resources. 

Here, we look at total deaths since the beginning of the pandemic. The story is the same for New York, New Jersey, and Massachusetts - with Michigan actively _distancing_ (ha) from them. Pennsylvania, Illinois, Indiana, and Maryland make a motley crew walking the middle path, while the rest seem to be following Washington's trajectory. This "rest" sees roughly 10 times fewer death per inhabitants than the "top" states.

### What about Europe?

```{r ec_incidence}
ec_incidence
```

It's easier to tell. They're all roughly behaving the same, apart from Sweden and the UK. The Covid-19 slowdown potentially induced by lockdowns is much more pronounced as shown by the sharp decrease in new cases over the last 20 days. 

```{r ec_deaths}
ec_deaths
```

Adjusting for population shows that all of the countries are roughly hanging out in the same ballpark with the exception of Germany, who either has a magic trick or is reporting deaths differently. 

Interestingly, we can see that France initially followed in the footsteps of Italy but has recently curbed deaths and will likely end up with fewer. This might be a result of having started strong lockdown earlier in their trajectory - as a result of Italy's early meltdown.

### And now, what we've all been waiting for...Europe vs. US

```{r us_eu_incidence}
us_eu_incidence
```

New York got hit later than European countries, but hard and fast. It's astounding. We can see that it's struggling to rein in the virus as fast as its US & European counterparts. 

California and Washington appear to be hanging out in a plateau, while Italy and France seem to have decidedly reduced the activity of the disease. 

```{r us_eu_deaths}
us_eu_deaths
```

The Dynamics are all overall similar, with California being hit surprisingly late in the game. The end levels, though are significantly different, with France and Italy almost exactly between the worst-hit (New-York) and best-behaving (Calif. & Wash.) states from the US.

Interestingly, we see that France & Italy report a higher level of mortality compared to California and Washington, but similar case counts. This illustrates differences in mesuring & reporting of Covid-19 cases and deaths.

### So what?

```{r us_eu_mortality, fig.width=10, echo = FALSE}
us_eu_mortality
```

Based on the downward trends that we see and assuming that there is no strong rebound, there is some hope that, on the 4th of July, the medical threat of Covid-19 will have greatly diminished.

On top of the uncertainty of the disease[^3], there is the uncertainty of international travel, border restrictions, and political context. The probability of COVID being under control on July 4 is high, but the probability that we don't have the legal right to party is quite high too. All in all, we couldn't gamble with our guests' nerves and risk a last-minute cancellation[^2].

In truth, we made the decision to postpone our wedding two weeks ago. As you can see, this was when we most of the states were continuing the climb to the top. Of course, we still believe postponement to be the right decision.[^4] Over these past couple of weeks, amidst our search for the sword that seals the darkness[^5], we have confirmed the changes with each of our vendors and made arrangements for the 14th of July 2021. 

[^3]: Antoine wanted to spend at least another 10 hours analyzing this. Isabelle would not allow it. 

[^2]: Two months before is not last-minute, right?

[^4]: Not to mention that we just found out that the _mairie_ (i.e. city hall) will not be allowing weddings until mid-July as of today. 

[^5]: This is a hint :) 

#### We can't wait to laugh and dance with all of you beautiful humans! 

