---
title: "Should we postpone our July 2020 wedding?"
author: "Isabelle Feldhaus, Antoine Lizée"
date: "2020-05-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("0. analysis.R")
```

With the World Health Organization's declaration of the Covid-19 pandemic in March 2020, the looming question has been whether or not to postpone our July wedding. A difficult decision before us, we wanted to take a look at the numbers to give some confidence to our decision, while keeping in mind that there are two key dimensions to consider for a just decision: (1) health and (2) geopolitics. 

We used data collated by the [New York Times](https://github.com/nytimes/covid-19-data/raw/master/us-states.csv) (yes, NYT appears to be the cleanest and most complete data source at the moment) and the [European CDC](https://opendata.ecdc.europa.eu/covid19/casedistribution/csv) from mid-February to May 3rd to get a picture of Covid-19 trajectories in the United States and selected European countries, respectively. We computed the cumulative cases and deaths over the course of the pandemic, incidence (cases per day), and mortality (deaths per day) and adjusted for population size for each state or country. 

Let's take a look. 

### The (Not So) United States

Indeed, there have been a lot of deaths in recent weeks...

```{r deaths_per_day, echo=FALSE}
plot(deaths_per_day)
```

This graph shows the number of deaths per day in the US. For comparison, at its peak, Covid-19 caused more than 2,000 fatalities per day when in normal times, we'd expect about 7,700 deaths each day on average. That's over 25%! 

Note the dips in the graph. These are weekends and, subsequently, pauses in data collection. Because of these idiosyncracies in reporting, all the following figures show values for the past 7 days. 

```{r us_incidence, echo=FALSE}
us_incidence
```

So, this graph is a mess. _But_ we see a few interesting trends. 

New York and New Jersey are moving together. Massachusetts is coming to get them. All that is unsurprising given their geographical proximity and interconnections. 

We also see a plateau. We can surmise that this mirrors the stay-home orders in various states. Good examples are Washington, New York, and Michigan (good luck finding this one). 

```{r us_deaths, echo=FALSE}
us_deaths
```

Looking at cases alone is not enough: case reporting is highly variable depending on testing protocols and resources. 

Here, we look at total deaths since the beginning of the pandemic. The story is the same for New York, New Jersey, and Massachusetts. Pennsylvania, Illinois, Indiana, and Maryland make a motley crew walking the middle path, while the rest seem to be following Washington's trajectory. 

### What about Europe?

```{r ec_incidence, echo=FALSE}
ec_incidence
```

It's easier to tell. They're all roughly behaving the same. The Covid-19 slowdown created by lockdowns is much more pronounced as shown by the sharp decrease in new cases over the last 20 days. 

```{r ec_deaths, echo=FALSE}
ec_deaths
```



```{r us_eu_incidence, echo=FALSE}
us_eu_incidence
```

```{r us_eu_mortality, echo=FALSE}
us_eu_mortality
```

```{r us_eu_deaths, echo=FALSE}
us_eu_deaths
```
